---
title: "AD Analysis Brendan"
output: html_document
date: "2025-04-08"
---

# Library Import
```{r}
# Library Import
library(BiocManager)
library(Seurat)
library(tidyverse)
library(patchwork) 
library(dittoSeq) 
library(stringr)
library(gprofiler2)
library(harmony)
library(cowplot)
library(openxlsx)
library(MAST)
library(devtools)
#library(escape)
home_dir <- "/mnt/vstor/SOM_PATH_DKB50/members/axb1812/AD_project"
gc()
Sys.setenv(TAR = "internal")
```



# Data Import
```{r}
counts <- readRDS(paste0(home_dir, "/GSE226602_rna_raw_counts.rds"))
```

```{r}
seurat_list <- CreateSeuratObject(counts = counts)
rm(counts)
seurat_list <- SplitObject(seurat_list, split.by = "orig.ident")
```

```{r}
samples <- names(seurat_list)
```

# Checking QC
```{r}
# Mitochondrial quality control
for(i in samples){ 
  seurat_list[[i]][["percent.mt"]] <- PercentageFeatureSet(seurat_list[[i]], pattern = "^MT-") 
} 

# RNA quality control
for(i in samples){ 
  seurat_list[[i]][["rna.size"]] <-   log10(Matrix::colSums(seurat_list[[i]]@assays$RNA$counts)) 
} 

```

```{r}
rna_size_min <- list() 
rna_size_max <- list() 

# Calculates min and max for rna filtering
for(i in samples){ 
  rna_size_min[[i]] <- median(seurat_list[[i]]@meta.data$rna.size) - (3*mad(seurat_list[[i]]@meta.data$rna.size)) 
  
  rna_size_max[[i]] <- median(seurat_list[[i]]@meta.data$rna.size) + (3*mad(seurat_list[[i]]@meta.data$rna.size)) 
} 
```

```{r}
# Visualizing quality control metrics of data
combined <- merge(x = seurat_list[[1]], y = seurat_list[2:length(seurat_list)], merge.data=TRUE)
# Violin Plot of rna size
p1 <- VlnPlot(combined,"rna.size", group.by = "orig.ident", raster = FALSE)
ggsave(filename= paste0(home_dir, "/rna_size.jpg"), width = 20, height = 20) # name the file as an EPS


# Violin Plot of mitochondrial content
p2 <- VlnPlot(combined,"percent.mt", group.by = "orig.ident", raster = FALSE)
ggsave(filename= paste0(home_dir, "/percent_mt.jpg"), width = 20, height = 20) # name the file as an EPS
rm(combined, p1, p2)
```

No QC necessary

```{r include=FALSE}
# S phase genes
hsp_s <- cc.genes.updated.2019$s.genes
# G2M phase genes
hsp_g2m <- cc.genes.updated.2019$g2m.genes

# Normalizes data and calculates cell-cycle scoring to be regressed
for(i in samples){ 
    seurat_list[[i]] <- NormalizeData(seurat_list[[i]]) 
    seurat_list[[i]] <- CellCycleScoring(seurat_list[[i]], s.features = hsp_s, g2m.features = hsp_g2m) 
}
```

# Running SCTransform
```{r include=FALSE}
options(future.globals.maxSize = 6000 * 1024^2)
#Applies SCTransform individually to each sample 

processed_list <- list()

total_initial <- length(seurat_list) + length(processed_list)

while (length(seurat_list) > 0) {
  cat("Remaining objects to process:", length(seurat_list), "of", total_initial, "\n")
  
  # Take the first object from the list
  obj <- seurat_list[[1]]
  
  # Run SCTransform on the object
  processed_obj <- SCTransform(
    obj,
    method = "glmGamPoi",
    return.only.var.genes = FALSE,
    vars.to.regress = c("S.Score", "G2M.Score"),
    verbose = FALSE
  )
  
  # Append the processed object to the processed_list
  processed_list[[length(processed_list) + 1]] <- processed_obj
  
  # Remove the first object from seurat_list
  seurat_list <- seurat_list[-1]
  
  cat("Processed objects so far:", length(processed_list), "\n")
}
```

```{r}
# Select integration features across samples
variable.features <- SelectIntegrationFeatures(object.list = processed_list, nfeatures = 3000)

# Start with the first Seurat object
combined <- processed_list[[1]]

# Create a list with the remaining objects
remaining_objects <- processed_list[-1]

# Optionally remove the original list to free memory
rm(processed_list)
gc()

# Merge objects sequentially while removing them from the list and printing progress
while (length(remaining_objects) > 0) {
  # Merge the first remaining object into 'combined'
  combined <- merge(combined, remaining_objects[[1]])
  
  combined <- JoinLayers(combined, assay = "RNA")

  # Remove the merged object from the list
  remaining_objects[[1]] <- NULL
  
  # Print the number of objects left to merge
  cat("Remaining objects to merge:", length(remaining_objects), "\n")
  
  # Prompt garbage collection to free memory
  gc()
}


# Assigns variable features calculated across samples
VariableFeatures(combined) <- variable.features 
```





```{r}
combined <- RunPCA(combined, verbose = FALSE)

p1 <- DimPlot(object = combined, reduction = "pca", pt.size = .1, group.by = "orig.ident") + NoLegend()

ggsave(filename= paste0(home_dir, "/harmony_pre_umap.jpg"), width = 7, height = 7) # name the file as an EPS

p2 <- VlnPlot(object = combined, features = "PC_1", group.by = "orig.ident", pt.size = 0) + NoLegend()

ggsave(filename= paste0(home_dir, "/harmony_pre_vln.jpg"), width = 7, height = 7) # name the file as an EPS

combined <- RunHarmony(combined, assay.use = "SCT", group.by.vars = "orig.ident", plot_convergence = FALSE) 

p1 <- DimPlot(object = combined, reduction = "harmony", pt.size = .1, group.by = "orig.ident") + NoLegend()

ggsave(filename= paste0(home_dir, "/harmony_post_umap.jpg"), width = 7, height = 7) # name the file as an EPS

p2 <- VlnPlot(object = combined, features = "harmony_1", group.by = "orig.ident", pt.size = 0) + NoLegend()

ggsave(filename= paste0(home_dir, "/harmony_post_vln.jpg"), width = 7, height = 7) # name the file as an EPS

p1 <- ElbowPlot(combined, reduction = "harmony", n = 50)

ggsave(filename= paste0(home_dir, "/elbowplot.jpg"), width = 7, height = 7) # name the file as an EPS
```

# Clustering
```{r}
#Performing Dimensional Reduction 
# UMAP
combined <- RunUMAP(combined, reduction = "harmony", dims = 1:25, verbose = FALSE) 

# Find Neighboring cells 
combined <- FindNeighbors(combined, reduction = "harmony", dims = 1:25, verbose = FALSE) 

# knn clustering
combined <- FindClusters(combined, resolution = 0.6, verbose = FALSE) 
```

```{r}
p1 <- dittoDimPlot(combined, "seurat_clusters", do.label = TRUE, legend.show = FALSE, labels.size = 3, raster = FALSE) 
ggsave(filename= paste0(home_dir, "/umap_clusters.jpg"), width = 10, height = 10) # name the file as an EPS

p1 <- dittoDimPlot(combined, "orig.ident",  raster = FALSE) + NoLegend()
ggsave(filename= paste0(home_dir, "/umap_orig.jpg"), width = 10, height = 10) # name the file as an EPS

p1 <- dittoDimPlot(combined, "percent.mt",  raster = FALSE) 
ggsave(filename= paste0(home_dir, "/umap_mt.jpg"), width = 10, height = 10) # name the file as an EPS
```


```{r}
#saveRDS(combined , paste0(home_dir, "/ad_data.RData"))
combined <- readRDS(paste0(home_dir, "/ad_data.RData"))
```


# Finding Sub Cluster for 14
```{r}
combined <- FindSubCluster(
  combined,
  14,
  graph.name = "SCT_snn", 
  subcluster.name = "sub.cluster",
  resolution = 0.1,
  algorithm = 1
)

Idents(combined) <- combined@meta.data$sub.cluster
```

```{r}
p1 <- dittoDimPlot(combined, "sub.cluster",  raster = FALSE) 
ggsave(filename= paste0(home_dir, "/umap_sub.cluster.jpg"), width = 10, height = 10) # name the file as an EPS
```



# Adding Metadata
```{r}
hc <- c("G1028", "G1273", "G991", "G1180", "G1279", "G598", "G780",
        "G1020", "G968", "G781", "G1111", "G912", "G978", "G1282",
        "G820", "G1162", "G989", "G1010", "G836", "G1200", "G970",
        "G905")

combined@meta.data$phenotype <- if_else(combined@meta.data$orig.ident %in% hc, "C", "AD")
```


# Finding Markers of Clusters
```{r}
# Get unique identifiers
unique_idents <- as.character(unique(combined@meta.data$sub.cluster))
unique_idents 


markers_list <- list()

# Loop over each identifier and run FindMarkers
for (id in unique_idents) {
  markers_list[[id]] <- FindMarkers(
    object = combined,
    ident.1 = id,
    only.pos = TRUE,
    assay = "RNA",
    max.cells.per.ident = 500
  )
}

saveRDS(markers_list, paste0(home_dir, "/ad_markers.RData"))
```


```{r}
markers_list <- readRDS(paste0(home_dir, "/ad_markers.RData"))
```


```{r}
names(markers_list)
names(markers_list) <- c("3 - CD8+ T", "7 -Low Quality T", "0 - CD4+ T", "2 - CD4+ T", "10 - NK", "1 - NK", "4 - CD4+ T", "8 - CD8+ T",
                         "11 - TRBV7-2+ CD4+ T", "13 - Non-classical Monocytes", "9 - Classical Monocyte", "5 - CD4+ T", "6 - B", "12 - NK",
                         "16 - Platelet", "15 - B", "14_1 - pDCs",
                         "14_0 - Erythroid", "14_2 - Progenitor")
```



0 - Ribosomal CD4+ T
1 - GNLY+ NK Cells
2 - T Cells (Naive Likely)
3 - CD8+ T
4 - T Cells (Naive Likely)
5 - Macs?
6 - B Cells
7  - Low Quality T
8 - CD8+ T
9 - Classical Monocyte
10 - NK Cells
11 - TRBV7-2+ T
12 - PTGDS+ NK Cells
13 - Non-Classical Monocytes
14_0 - Erythroid
14_1 - pDCs
14_2 - Progenitor
15 - B Cells
16 - Platelet



```{r}
# Summary Idents

new_levels <- c("0", "1", "2","3" ,"4", "5", "6", "7", "8", "9", "10",
                "11", "12", "13", "14_0", "14_1", "14_2", "15", "16")

Idents(combined) <- combined@meta.data$sub.cluster 

Idents(combined) <- factor(Idents(combined), levels = new_levels)

new.cluster.ids <- c("CD4+ T", "NK", "CD4+ T", "CD8+ T", "CD4+ T",
                     "CD4+ T", "B", "Low Quality", "CD8+ T", "Classical Monocyte",
                     "NK", "TRB7-2+ T", "NK", "Non-Classical Monocytes", "Erythroid", "pDCs",
                     "Progenitor", "B", "Platelet")

names(new.cluster.ids) <- levels(combined) 

combined <- RenameIdents(combined, new.cluster.ids) 

combined@meta.data$summary <- Idents(combined) 
```

```{r}
p1 <- dittoDimPlot(combined, "summary", raster = FALSE) + NoLegend()
ggsave(filename= paste0(home_dir, "/umap_summary.svg"), width = 10, height = 10) # name the file as an EPS
```

```{r}
saveRDS(combined , paste0(home_dir, "/ad_data.RData"))
```



# Differential Expression by Phenotype
```{r}
unique_celltypes <- unique(combined@meta.data$summary)

marker_list <- list()

for(i in unique_celltypes){
seurat_obj <- subset(combined, subset = summary == i)

Idents(seurat_obj) <- seurat_obj@meta.data$phenotype

seurat_obj <- NormalizeData(seurat_obj)
temp <- FindMarkers(seurat_obj, ident.1 = "C", ident.2 = "AD", only.pos = FALSE, assay = "RNA", logfc.threshold = 0.5, test.use = "MAST", latent.vars = "orig.ident")

marker_list[[i]] <- temp
print(paste0("Finished:", i))
}

saveRDS(marker_list, file = paste0(home_dir, "/de_condition_AD.RData"))
```


```{r}
saveRDS(full_list, file = paste0(home_dir, "/de_condition_AD.RData"))
```



# Adding Module score
```{r}
Q_T2D_GSE <- read.csv(file = paste0(home_dir, "/Q_T2D.csv"))
# Prepare the Q matrix with only PC2
Q_T2D_GSE <- as.data.frame(Q_T2D_GSE)
Q_T2D_GSE <- dplyr::select(Q_T2D_GSE, "GeneID", "PC2")
# Get the top 50 PC scores
top_50_PC2 <- Q_T2D_GSE %>%
  arrange(desc(PC2)) %>%
  slice_head(n = 50)
# Get the bottom 50 PC scores
bottom_50_PC2 <- Q_T2D_GSE %>%
  arrange(PC2) %>%
  slice_head(n = 50)
# Combine the top 50 and bottom 50 PC scores into a single data frame
Q_T2D_PC2 <- bind_rows(top_50_PC2, bottom_50_PC2)
# Keep gene list that match from the top 50 and bottom 50 loadings in T2D PC2
TopBotPC2genelist <- Q_T2D_PC2$GeneID
```

```{r}
TopBotPC2genelist <- intersect(TopBotPC2genelist, rownames(combined))
TopBotPC2genelist <- list(TopBotPC2genelist) # 90 Genes left after filtering for AddModuleScore
```

```{r}
combined <- AddModuleScore(combined, features = TopBotPC2genelist, name = "PC2_score_final")
```


```{r}
p1 <- dittoDimPlot(combined, "PC2_score_final1", raster = FALSE, max.color = "#F0E442", min.color = "#0072B2") + NoLegend()
ggsave(filename= paste0(home_dir, "/umap_PC2_score.svg"), width = 10, height = 10) # name the file as an EPS
```


```{r}
df <- data.frame(combined@meta.data$summary, combined@meta.data$PC2_score_final1)
names(df) <- c("summary_type", "PC2_score")
write.csv(df, paste0(home_dir, "/modulescore_df.csv"))

```



# Filtering DE for presence of PC2 Genes
```{r}
unique_celltypes <- names(full_list)

de_list <- list()

for(i in unique_celltypes){
  temp <- full_list[[i]]
  temp <- temp %>%
    filter(row.names(temp) %in% TopBotPC2genelist[[1]])
  de_list[[i]] <- temp
}
```


```{r}
saveRDS(de_list, file = paste0(home_dir, "/de_condition_pc2_filtered.RData"))
```

# Scoring cells by genes present
```{r}
genes_to_score <- TopBotPC2genelist[[1]]

rna_counts <- GetAssayData(object = combined, assay = "RNA", slot = "counts")

common_genes <- intersect(genes_to_score, rownames(rna_counts))
if (length(common_genes) == 0) {
  stop("None of the genes in your list are present in the Seurat object!")
} else {
  message("Found ", length(common_genes), " genes in common out of ", length(genes_to_score))
}

# For each cell, count how many of the genes in the common list have counts > 0 (i.e., expressed)
gene_expression_score <- colSums(rna_counts[common_genes, ] > 0)

# Add the count as a new metadata column 'gene_present'
combined <- AddMetaData(object = combined, metadata = gene_expression_score, col.name = "gene_present")

# Optional: print the head of the metadata to verify the new column
head(combined@meta.data)
```

```{r}
p1 <- dittoDimPlot(combined, "gene_present", raster = FALSE, max.color = "#F0E442", min.color = "#0072B2") + NoLegend()
ggsave(filename= paste0(home_dir, "/umap_gene_found.svg"), width = 10, height = 10) # name the file as an EPS
```


```{r}
sessionInfo()
```

